Efficient simulation of individual-based population models

### 2024-03-18 article.qmd
update

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université, CNRS, Institut du Risque et de l'Assurance 
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
date: last-modified
date-modified: last-modified
description: |
    TODO
abstract: >+
    The `R` Package `IBMPopSim` aims to simulate the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: default
  computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: false
    echo: true
---

@@

In @sec-IBM, we introduce the mathematical framework that characterizes the class of Stochastic Individual-Based Models (IBMs) that can be implemented in the `IBMPopSim` package. In particular, a general pathwise representation of IBMs is presented. The population dynamics is obtained as the solution of an SDE driven by Poisson measures, for which we obtain existence and uniqueness results in Theorem [1](#ThEqZ){reference-type="ref" reference="ThEqZ"}. Additionally, a succinct overview of the package is provided. In @sec-simulation the two main algorithms for simulating the population evolution of an IBM across the interval $[0, T]$ are detailed. In @sec-package we present the main functions of the `IBMPopSim` package, which allow for the definition of events and their intensities, the creation of a model, and the simulation of scenarios. Two examples are detailed in @sec-insurance-portfolio and sec-example-interaction, featuring applications involving an heterogeneous insurance portfolio characterized by entry and exit events, and an age and size-structured population with intricate interactions.

# Stochastic Individual-Based Models (IBMs) in IBMPopSim {#sec-IBM}

@@

### Model creation 
To implement this model in IBMPopSim, it is necessary to individually define each event type. In this example, the `mk_event_individual` function is used. The creation of an event involves  a few lines of `cpp` instructions defining the intensity and, if applicable, the kernel of the event. For a more in depth description of the event creation step and its parameters, we refer to @sec-package_events.

@@

To summarize, an individual intensity in IBMPopSim can be written as:
$$
    \lambda^e_t(I,Z_t) = \lambda^e(t, I) \mathbf{1}_{\{e \in \mathcal{E}\}} + \biggl( \sum_{j=1}^{N_t} W^e(t, I, I_j) \biggr) \mathbf{1}_{\{e \in \mathcal{E}_W\}}.
$$ {#eq-indiviual-intensity}

@@

The proof of @thm-eq-Z is detailed in the Appendix @prf-thm-eq-Z. Note that @eq-sde-eq-Z is an SDE describing the evolution of the IBM, the intensity of the events in the right hand side of the equation depending on the population process $Z$ itself. 
The main idea of the proof of @thm-eq-Z is to use the non explosion property of @lem-non-explosion, and to write the r.h.s of @eq-sde-eq-Z as a sum of simple equations between two successive events, solved by induction.

The proof of @lem-non-explosion
, detailed in Appendix @prf-lem-non-explosion
is more technical and rely on pathwise
comparison result, generalizing those obtained in [@KaaElK20]. An
alternative pathwise representation of the population process, inspired
by the randomized Algorithm
[\[algo::rzndomized\]](#algo::rzndomized){reference-type="ref"
reference="algo::rzndomized"}, 
is given as well in Theorem
[\[ThEqZrandomized\]](#ThEqZrandomized){reference-type="ref"
reference="ThEqZrandomized"}.

::: {#thm-eq-Z} 
## Pathwise representation

Let $T\in {\mathbb{R}}^+$ and $\mathcal{J}= \mathbb N \times(\mathcal E \cup \mathcal{E}_W)$. 
Let $Q$ be a random Poisson measure on $\mathbb R^+ \times \mathcal{J}\times \mathbb{R}^+$, of intensity $\mathrm{d}t \delta_{\mathcal{J}}(\mathrm{d}k,\mathrm{d}e)  \mathbf{1}_{[0,\bar \lambda^e]} (\theta)\mathrm{d}\theta$, with $\delta_{\mathcal{J}}$ the counting measure on $\mathcal{J}$. 
Finally, let $Q^{\mathcal P}$ be a random Poisson measure on $\mathbb R^+ \times \mathcal{P}  \times \mathbb{R}^+$, of intensity $\mathrm{d}t \delta_{\cal P}(\mathrm{d}e)  \mathbf{1}_{[0,\bar \mu^e]} (\theta)\mathrm{d}\theta$, and $Z_0= \sum_{k=1}^{N_0} \delta_{I_k}$ an initial population.
Then, under @cnj-non-explosion
, there exists a unique
measure-valued population process $Z$, strong solution on the following
SDE driven by the Poisson measure $Q$: 
$$
    \begin{aligned}
    Z_t = Z_0 &+ \int_0^t \int_{ \mathcal{J}\times \mathbb R^+ }\phi^e (s , I_k)  \mathbf{1}_{\{k \leq N_{s^-}\} }\mathbf{1}_{\{\theta \leq \lambda_s^e(I_k, Z_{s^-})\}} Q (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ) \\
    \nonumber &+   \int_0^t \int_{\mathcal{P} \times \mathbb R^+}  \phi^e(s, I_{s^-}) \mathbf{1}_{\{\theta \leq \mu^e(s) \}} Q^{\mathcal{P}} (\mathrm{d}s ,\mathrm{d}e,  \mathrm{d}\theta),  \qquad \forall  0 \leq t \leq T,
\end{aligned}
$$ {#eq-sde-eq-Z}
and where $I_{s^-}$ is an individual, chosen uniformly among alive individuals in the population $Z_{s^-}$.
:::

@@

@@

Let us now come back to the simulation of the IBM introduced in @sec-ibm. For ease of notations, we assume that there are no event with Poisson intensity ($\mathcal{P} =\emptyset$), so that all events that occur are of type $e \in \mathcal{E} \cup \mathcal{E}_W$, with individual intensity $\lambda_t^e(I,Z_t)$ depending on the population composition $Z_t$ ($e \in \mathcal{E}_W$) or not ($e \in \mathcal{E}$), as defined in @eq-indivual-intensity and verifying either @cnj-intensity1 or @cnj-intensity2.
The global intensity @eq-global-event-intensity at time $t \in [0,T]$ is thus
$$
    \Lambda_t(Z_t) = \sum_{e \in \mathcal{E}} \Big( \sum_{k=1}^{N_t} \lambda^e(t, I_k) \Big)
    + \sum_{e \in \mathcal{E}_W} \Big( \sum_{k=1}^{N_t} \sum_{j=1}^{N_t} W^e(t, I_k, I_j) \Big) \leq \bar \Lambda(N_t),
$$ {#eq-def-intensity}
with
$\bar \Lambda(n) = \big(\sum_{e \in \mathcal{E}} \bar \lambda^e \big) n + \big( \sum_{e \in \mathcal{E}_W} \bar W^e \big) n^2$.

@@

@@

### Proof of @thm-eq-Z

::: prf
For ease of notation, we prove the case when $\mathcal{P} =\emptyset$ (there are no events with Poisson intensity).

**Step 1** The existence of a solution to @eq-sde-eq-Z is obtained by induction. Let $Z^1$ be the unique solution the thinning equation:
$$
    Z_t^1 = Z_0 + \int_0^t \int_{ \mathcal{J}\times \mathbb R^+ }\phi^e (s , I_k)  \mathbf{1}_{\{k \leq N_{0}\} }\mathbf{1}_{\{\theta \leq \lambda_s^e(I_k, Z_{0})\}} Q (\mathrm{d}s,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ), \quad \forall  0 \leq t \leq T.
$$
Let $T_1$ be the first jump time of $Z^1$. Since $Z_{s^-}^1 = Z_{0}$ and $N_{s^-}=N_{0}$ on $[0, T_1]$, $Z^1$ is solution of @eq-sde-eq-Z on $[0,T_1]$.

@@

**Step 2** Let $\tilde Z$ be a solution of @eq-sde-eq-Z.
Using the same arguments than in Step 1, it is straight forward to show
that $\tilde Z$ coincides with $Z^n$ on $[0,T_n]$, for all $n \geq 1$.
Thus, $\tilde{Z} = Z$, with achieves to prove uniqueness.
:::

>>>>

---
title: "Efficient simulation of individual-based population models"
subtitle: "The `R` package `IBMPopSim`"
author:
  - name: Daphné Giorgi 
    email: daphne.giorgi@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~giorgi/
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Vincent Lemaire 
    corresponding: true
    email: vincent.lemaire@sorbonne-universite.fr
    url: https://perso.lpsm.paris/~vlemaire/
    orcid: 0000-0002-0433-7722
    affiliations:
      - name: Sorbonne Université, CNRS
        department: Laboratoire de Probabilités, Statistique et Modélisation
        url: https://www.lpsm.paris/
  - name: Sarah Kaakai
    corresponding: true
    email: sarah.kaakai@univ-lemans.fr
    url: https://perso.univ-lemans.fr/~skaakai/
    affiliations:
      - name: Le Mans Université, CNRS, Institut du Risque et de l'Assurance 
        department: Laboratoire Manceau de Mathématiques
        url: https://lmm.univ-lemans.fr/fr/
date: last-modified
date-modified: last-modified
description: |
    TODO
abstract: >+
    The `R` Package `IBMPopSim` aims to simulate the random evolution of heterogeneous populations using stochastic Individual-Based Models (IBMs). The package enables users to simulate population evolution, in which individuals are characterized by their age and some characteristics, and the population is modified by different types of events, including births/arrivals, death/exit events, or changes of characteristics. The frequency at which an event can occur to an individual can depend on their age and characteristics, but also on the characteristics of other individuals (interactions). Such models have a wide range of applications in fields including  actuarial science, biology, ecology or epidemiology. `IBMPopSim` overcomes the limitations of time-consuming IBMs simulations by implementing new efficient algorithms  based on thinning methods, which are compiled using the `Rcpp` package while providing a user-friendly interface.
keywords: [Individual-based models, stochastic simulation, population dynamics, Poisson measures, thinning method, actuarial science, insurance portfolio simulation]
citation:
  type: article-journal
  container-title: "Computo"
  doi: "xxxx"
  url: https://computo.sfds.asso.fr/template-computo-quarto
  publisher: "Société Française de Statistique"
  issn: "2824-7795"
bibliography: references.bib
github-user: VincentLemaire
repo: "ibmpopsim_computo"
draft: true # set to false once the build is running
published: false # will be set to true once accepted
format:
  computo-html: default
  computo-pdf: default
crossref:
    cnj-title: Assumption
    cnj-prefix: Assumption 
execute:
    eval: true
    echo: true
    warning: false
---

@@

In @sec-ibm, we introduce the mathematical framework that characterizes the class of Stochastic Individual-Based Models (IBMs) that can be implemented in the `IBMPopSim` package. In particular, a general pathwise representation of IBMs is presented. The population dynamics is obtained as the solution of an SDE driven by Poisson measures, for which we obtain existence and uniqueness results in @thm-eq-Z. Additionally, a succinct overview of the package is provided. 
In @sec-simulation the two main algorithms for simulating the population evolution of an IBM across the interval $[0, T]$ are detailed. 
In @sec-package we present the main functions of the `IBMPopSim` package, which allow for the definition of events and their intensities, the creation of a model, and the simulation of scenarios. 
Two examples are detailed in @sec-insurance-portfolio and sec-example-interaction, featuring applications involving an heterogeneous insurance portfolio characterized by entry and exit events, and an age and size-structured population with intricate interactions.

# Stochastic Individual-Based Models (IBMs) in IBMPopSim {#sec-ibm}

@@

### Model creation 
To implement this model in IBMPopSim, it is necessary to individually define each event type. In this example, the `mk_event_individual` function is used. The creation of an event involves  a few lines of `cpp` instructions defining the intensity and, if applicable, the kernel of the event. For a more in depth description of the event creation step and its parameters, we refer to @sec-package-events.

@@

To summarize, an individual intensity in IBMPopSim can be written as:
$$
    \lambda^e_t(I,Z_t) = \lambda^e(t, I) \mathbf{1}_{\{e \in \mathcal{E}\}} + \biggl( \sum_{j=1}^{N_t} W^e(t, I, I_j) \biggr) \mathbf{1}_{\{e \in \mathcal{E}_W\}}.
$$ {#eq-individual-intensity}

@@

::: {#thm-eq-Z} 
## Pathwise representation
Let $T\in {\mathbb{R}}^+$ and $\mathcal{J}= \mathbb N \times(\mathcal E \cup \mathcal{E}_W)$. 
Let $Q$ be a random Poisson measure on $\mathbb R^+ \times \mathcal{J}\times \mathbb{R}^+$, of intensity $\mathrm{d}t \delta_{\mathcal{J}}(\mathrm{d}k,\mathrm{d}e)  \mathbf{1}_{[0,\bar \lambda^e]} (\theta)\mathrm{d}\theta$, with $\delta_{\mathcal{J}}$ the counting measure on $\mathcal{J}$. 
Finally, let $Q^{\mathcal P}$ be a random Poisson measure on $\mathbb R^+ \times \mathcal{P}  \times \mathbb{R}^+$, of intensity $\mathrm{d}t \delta_{\cal P}(\mathrm{d}e)  \mathbf{1}_{[0,\bar \mu^e]} (\theta)\mathrm{d}\theta$, and $Z_0= \sum_{k=1}^{N_0} \delta_{I_k}$ an initial population.
Then, under @cnj-non-explosion
, there exists a unique
measure-valued population process $Z$, strong solution on the following
SDE driven by the Poisson measure $Q$: 
$$
    \begin{aligned}
    Z_t = Z_0 &+ \int_0^t \int_{ \mathcal{J}\times \mathbb R^+ }\phi^e (s , I_k)  \mathbf{1}_{\{k \leq N_{s^-}\} }\mathbf{1}_{\{\theta \leq \lambda_s^e(I_k, Z_{s^-})\}} Q (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ) \\
    \nonumber &+   \int_0^t \int_{\mathcal{P} \times \mathbb R^+}  \phi^e(s, I_{s^-}) \mathbf{1}_{\{\theta \leq \mu^e(s) \}} Q^{\mathcal{P}} (\mathrm{d}s ,\mathrm{d}e,  \mathrm{d}\theta),  \qquad \forall  0 \leq t \leq T,
\end{aligned}
$$ {#eq-sde-eq-Z}
and where $I_{s^-}$ is an individual, chosen uniformly among alive individuals in the population $Z_{s^-}$.
:::

The proof of @thm-eq-Z is detailed in the Appendix, @sec-prf-thm-eq-Z. Note that @eq-sde-eq-Z is an SDE describing the evolution of the IBM, the intensity of the events in the right hand side of the equation depending on the population process $Z$ itself. 
The main idea of the proof of @thm-eq-Z is to use the non explosion property of @lem-non-explosion, and to write the r.h.s of @eq-sde-eq-Z as a sum of simple equations between two successive events, solved by induction.

@@

The proof of @lem-non-explosion, detailed in Appendix @sec-prf-lem-non-explosion is more technical and rely on pathwise comparison result, generalizing those obtained in [@KaaElK20]. An alternative pathwise representation of the population process, inspired by the randomized @cnj-randomized is given as well in @thm-randomized.

@@

Let us now come back to the simulation of the IBM introduced in @sec-ibm. For ease of notations, we assume that there are no event with Poisson intensity ($\mathcal{P} =\emptyset$), so that all events that occur are of type $e \in \mathcal{E} \cup \mathcal{E}_W$, with individual intensity $\lambda_t^e(I,Z_t)$ depending on the population composition $Z_t$ ($e \in \mathcal{E}_W$) or not ($e \in \mathcal{E}$), as defined in @eq-individual-intensity and verifying either @cnj-intensity1 or @cnj-intensity2.
The global intensity @eq-global-ev-intensity at time $t \in [0,T]$ is thus
$$
    \Lambda_t(Z_t) = \sum_{e \in \mathcal{E}} \Big( \sum_{k=1}^{N_t} \lambda^e(t, I_k) \Big)
    + \sum_{e \in \mathcal{E}_W} \Big( \sum_{k=1}^{N_t} \sum_{j=1}^{N_t} W^e(t, I_k, I_j) \Big) \leq \bar \Lambda(N_t),
$$ {#eq-def-intensity}
with
$\bar \Lambda(n) = \big(\sum_{e \in \mathcal{E}} \bar \lambda^e \big) n + \big( \sum_{e \in \mathcal{E}_W} \bar W^e \big) n^2$.

@@

TODO

# Model creation and simulation with IBMPopSim {#sec-package}

The use of the `IBMPopSim` package is mainly done in two steps: a first model creation followed by the simulation of the population evolution. The creation of a model is itself based on two steps: the description of the population $Z_t$, as introduced in @sec-population, and the description of the events types, along with their associated intensities, as detailed in @sec-events and @sec-event-intensity.
A model is compiled by calling the `mk_model` function, which internally uses a template mechanism to generate automatically the source code describing the model, which is subsequently compiled using the `Rcpp` package to produce the object code.

After the compilation of the model, the simulations are launched by calling the `popsim` function. This function depends on the previously compiled model and simulates a random trajectory of the population evolution based on an initial population and on parameter values, which can change from a call to another.

In this section, we take a closer look at each component of a model in `IBMPopSim`. We also refer to the [IBMPopSim website](https://daphnegiorgi.github.io/IBMPopSim/) and to the `vignettes` of the package for more details on the package and various examples of model creation.

A population $Z$ is represented by an object of class `population` containing a data frame where each row corresponds to an individual $I=(\tau^b, \tau^d, x)$, and which has at least two columns, `birth` and `death`, corresponding to the birth date $\tau^b$ and death/exit date $\tau^d$ ($\tau^d$ is set to `NA` for alive individuals). The data frame can contain more than two columns if individuals are described by additional characteristics $x= (x_1,\dots x_n)$.

If entry events can occur in the population, the population shall contain a characteristic named `entry`. This can be done by setting the flag `entry=TRUE` in the `population` function, or by calling the `add_characteristic` function on an existing population. During the simulation, the date at which an individual enters the population is automatically recorded in the variable `I.entry`.
If exit events can occur, the population shall contain a characteristic named `out`. This can be done by setting the flag `out=TRUE` in the `population` function, or by calling the `add_characteristic` function. When an individual `I` exits the population during the simulation, `I.out` is set to `TRUE` and its exit time is recorded as a "death" date.

In the example below, individuals are described by their birth and death dates, as well a Boolean characteristics called male, and the `entry` characteristic. For instance, the first individual is a female whose age at $t_0=0$ is $107$ and who was originally in the population.
```{R}
pop_init <- population(EW_pop_14$sample, entry=TRUE)
str(pop_init)
```

### Individual

In the `C++` model which is automatically generated and compiled, an individual `I` is an object of an internal class containing some attributes (`birth_date`, `death_date` and the characteristics), and some methods including:

- `I.age(t)`: a `const` method returning the age of an individual `I` at time `t`,
- `I.set_age(a, t)`: a method to set the age `a` at time `t` of an individual `I` (set `birth_date` at `t-a`),
- `I.is_dead(t)`: a `const` method returning `true` if the individual `I` is dead at time `t`.

::: {#rem-characteristics}
A characteristic $x_i$ must be of atomic type: `logical`, `integer`, `double` or `character`. The function `get_characteristic` allows to easily get characteristics names and their types from a population data frame. We draw the attention to the fact that some names for characteristics are forbidden, or reserved to specific cases : this is the case for `birth`, `death`, `entry`, `out`, `id`.
:::

## Events {#sec-package-events}

The most important step of the model creation is the events creation. The call to the function creating an event is of form
```{R}
#| eval: false 
mk_event_CLASS(type="TYPE", name="NAME", ...)
```
where `CLASS` is replaced by the class of the event intensity, described in @sec-event-intensity , and `type` corresponds to the event type, described in Section @sec-events.
@tbl-intensity-classes and @tbl-event-types summarize the different possible choices for intensity classes and types of event. The optional argument `name` gives a name to the event. If not specified, the name of the event is its type, for instance `death`. However, a name must be specified if the model is composed of several events with the same type (for instance when there are multiple death events corresponding to different causes of death). The other arguments depend on the intensity class and on the event type.

::: {#tbl-panel layout-ncol=2}

Intensity class         Set               `CLASS`
----------------------- ----------------- -------------------------
Individual              $\mathcal{E}$     `individual`
Interaction             $\mathcal{E}_W$   `interaction`
Poisson                 $\mathcal{P}$     `poisson`
Inhomogeneous Poisson   $\mathcal{P}$     `inhomogeneous_poisson`

: Intensity Classes {#tbl-intensity-classes}

Event type   `TYPE`
------------ ---------
Birth        `birth`
Death        `death`
Entry        `entry`
Exit         `exit`
Swap         `swap`

: Event Types {#tbl-event-types}

Main caption
:::

The intensity function and the kernel of an event are defined through arguments of the function `mk_event_CLASS`. These arguments are strings composed of few lines of code. Since the model is compiled using `Rcpp`, the code should be written in `C++`. However, thanks to the functions/variables of the package, even the non-experienced `C++` user can define a model quite easily. 
To facilitate the implementation, the user can also define a list of **model parameters**, which can be used in the event and intensity definitions. These parameters are stored in a named list and can be of various types: atomic type, numeric vector or matrix, predefined function of one variable (`stepfun`, `linfun`, `gompertz`, `weibull`, `piecewise_x`), piecewise functions of two variables (`piecewise_xy`). 
We refer to the `vignette(IBMPopSim_cpp)` for more details on parameters types and basic `C++` tools. Another advantage of the model parameters is that their value can be modified from a simulation to another without changing the model.

# Insurance portfolio {#sec-insurance-portfolio}

TODO

# Population with genetically variable traits {#sec-example-interaction}

TODO

@@

### Proof of @thm-eq-Z {#sec-prf-thm-eq-Z}

::: {.proof}
For ease of notation, we prove the case when $\mathcal{P} =\emptyset$ (there are no events with Poisson intensity).

- Step 1. The existence of a solution to @eq-sde-eq-Z is obtained by induction. Let $Z^1$ be the unique solution the thinning equation:
$$
    Z_t^1 = Z_0 + \int_0^t \int_{ \mathcal{J}\times \mathbb R^+ }\phi^e (s , I_k)  \mathbf{1}_{\{k \leq N_{0}\} }\mathbf{1}_{\{\theta \leq \lambda_s^e(I_k, Z_{0})\}} Q (\mathrm{d}s,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ), \quad \forall  0 \leq t \leq T.
$$
Let $T_1$ be the first jump time of $Z^1$. Since $Z_{s^-}^1 = Z_{0}$ and $N_{s^-}=N_{0}$ on $[0, T_1]$, $Z^1$ is solution of @eq-sde-eq-Z on $[0,T_1]$.

@@

- Step 2. Let $\tilde Z$ be a solution of @eq-sde-eq-Z. Using the same arguments than in Step 1, it is straight forward to show that $\tilde Z$ coincides with $Z^n$ on $[0,T_n]$, for all $n \geq 1$. Thus, $\tilde{Z} = Z$, with achieves to prove uniqueness.
:::
